Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-06-01 writeup/pnas-word-order-si.tex
SI

@@

We base our formalization on the comprehensive study by Dryer \cite{dryer1992greenbergian}.\footnote{Regarding objections by \citet{dunn2011evolved}, we refer to the follow-ups by \citet{levy2011computational}, and \citet{croft2011greenbergian}.}
Greenberg's original study was based on 30 languages; more recently, Dryer \cite{dryer1992greenbergian} documented the word order correlations based on typological data from 625 languages.
\cite{dryer1992greenbergian} formulated these universals as correlations between the order of objects and verbs and the orders of other syntactic relations.
We test our ordering grammars for these correlations by testing whether the coefficients for these syntactic relations have the same sign as the coefficient of the verb-object relation.
Testing correlations is therefore constrained by the degree to which these relations are annotated in UD.
The verb-object relation corresponds to the  \emph{obj} relation defined by UD.
While most of the other relations also correspond to UD relations, some are not annotated reliably.
We were able formalize eleven out of Dryer's seventeen correlations in UD.
Six of these could not be expressed individually in UD, and were collapsed into three coarse-grained correlations:
First, tense/aspect and negative auxiliaries are together represented by the \emph{aux} relation in UD.
Second, the relation between complementizers and adverbial subordinators with their complement clauses is represented by the \emph{mark} relation.
Third, both the verb-PP relation and the relation between adjectives and their standard of comparison is captured by the \emph{obl} relation.

@@

Here we discuss how our formalization of communicative efficiency relates to formalizations that have been proposed in the information-theoretic literature on language.
Across the literature, the core idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms, with some differences in the precise formalization of these two quantities \cite{ferreri2003least,ferrericancho2007global, kemp2012kinship,xu2014numeral, regier2015word,xu2016historical, zaslavsky2018efficient,futrell2017memory,peloquin2019interactions, zaslavsky2019semantic, frank2012predicting,goodman2013knowledge, kao2014nonliteral, bennett2018extremely}.

@@

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parseability.
While the computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$, we also examined word order predictions for grammars optimized for only parseability or only predictability, in order to tease apart predictions made by these two components.
As shown in Table~\ref{tab:all-predictions-1}, each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions.
%Indeed, the same holds for all of the 38 syntactic relations in UD  (Table~\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})
%We additionally report results with $\lambda = 0$, i.e., placing all weight on parseability.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.\footnote{Results from one of the preliminary experiments reported in Figure~\ref{table:corr-resu-previous} show that results are stable to small variation of $\lambda$: Essentially equivalent predictions are obtained for $\lambda=1.0$. While $\lambda=1.0$ is not a valid choice for communicative efficiency in general due to the possibility of collapse to a single utterance, our setting does not allow such a collapse, as the syntactic structure already determines which words are present in the sentence.}
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}), in which case not all correlations are predicted (Figure~\ref{fig:posterior}).
%Finally, the robustness of our results to small variations of $\lambda$ is confime by reporting results with $\lambda=1$ in Table~\ref{table:corr-resu-previous}.

@@

\begin{table}
\small{
\begin{center}
\begin{tabular}{l||ll|ll|ll|ll|ll|ll|llllll}
%	   for(dep in c("obj", "lifted_mark", "acl", "nmod", "obl")) {
		   Language 
		   &	\multicolumn{2}{|c|}{Objects} 
		   &	\multicolumn{2}{|c|}{Adpositions} 
		   &	\multicolumn{2}{|c|}{Compl.} 
		   &	\multicolumn{2}{|c|}{Rel.Cl.} 
		   &	\multicolumn{2}{|c|}{Genitive} 
		   &	\multicolumn{2}{|c|}{PP}  \\ \hline\hline
	\input{../results/grammars/comparison-table.tex}
\end{tabular}
\end{center}
}
\caption{Comparing grammars extracted from databases to linguistic judgments in the World Atlas of Linguistic Structures. For each of the six syntactic relation, the first column provides the ordered coded in the extracted grammar; the second column provides the order coded in WALS (DH for dependent-head, HD for head-dependent order). `?' indicates that WALS has no data.
$*$ indicates that WALS does not list a dominant order, which can mean that both orders occur at similar rates in the language, or that insufficient data was available when compiling WALS \citep{dryer2011evidence}.
Finally, `--' indicates that the relation does not occur in the corpus.
}\label{tab:grammars-wals}
\end{table}

@@

Recall the definition of efficiency
\begin{equation}\label{eq:efficiency-recall}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where
\begin{equation}
	R_{Pars} := I[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)} 
\end{equation}
\begin{equation}
	R_{Pred} := - H[\utterance] = \sum_{u} p(u) \log p(u)
\end{equation}
where $t \sim \tree$ is a distribution over syntactic structures, and $u \sim p_{\mathcal{L}_\theta}(u|t)$ denotes the corresponding linearized sentences.

@@

The degree to which a parser with parameters $\phi$ succeeds in parsing a sentence $u$ with structure $t$ is\footnote{Note that, in the definition of $R_{Pars}$, the term $p(t)$ is a constant independent of $\phi$ and the word order grammar $\mathcal{L}_\theta$; it can therefore be ignored in the optimization process.} 

@@

	R_{Pars}^{\phi}(u,t) =  \log \frac{p_\phi(t|u)}{p(t)}

@@

Using this, we can rewrite Efficiency~(\ref{eq:efficiency-recall}), for a given grammar $\theta$, equivalently into the parseability and predictability achieved with the best parser and language models:

@@

	R_{\textit{Eff}}^{\theta} := \max_{\phi,\psi} R_{\textit{Eff}}^{\theta, \phi, \psi} := \max_{\phi,\psi} \E_t \E_{u \sim p_\theta(u|t)} \left[R_{Pars}^{\phi}(u,t) + \lambda R_{Pred}^{\psi}(u)\right]

@@

@@

In order to find an optimal grammar $\theta$, we thus need to solve

@@

\argmax_\theta\	\max_{\phi, \psi} R_{\textit{Eff}}^{\theta, \phi, \psi} %= \argmax_\theta\	\max_{\phi, \psi} \E_t \E_{u \sim p_\theta(u|t)} \left[R_{Pars}^{\phi}(u,t) + \lambda R_{Pred}^{\psi}(u)\right]

@@

where \ref{eq:dtheta} is derived using the \emph{score-function} or \emph{REINFORCE} trick~\cite{williams1992simple}.

@@

We can therefore apply stochastic gradient descent to jointly optimize $\theta, \phi, \psi$:
In each optimization step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim p_{\theta}(\cdot|t)$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator given by the expressions in the square brackets in \ref{eq:dtheta}-\ref{eq:dpsi}.

>>>>

\textcolor{red}{TODOs}

@@

We base our formalization on the comprehensive study by Dryer \cite{dryer1992greenbergian}.\footnote{Regarding objections by \citet{dunn2011evolved}, we refer to the follow-ups by \citet{levy2011computational}, and \citet{croft2011greenbergian}.}
Greenberg's original study was based on 30 languages; more recently, Dryer \cite{dryer1992greenbergian} documented the word order correlations based on typological data from 625 languages.
\citet{dryer1992greenbergian} formulated these universals as correlations between the order of objects and verbs and the orders of other syntactic relations.
We test our ordering grammars for these correlations by testing whether the coefficients for these syntactic relations have the same sign as the coefficient of the verb-object relation.
Testing correlations is therefore constrained by the degree to which these relations are annotated in UD.
The verb-object relation corresponds to the  \emph{obj} relation defined by UD.
While most of the other relations also correspond to UD relations, some are not annotated reliably.
We were able formalize eleven out of Dryer's seventeen correlations in UD.
Six of these could not be expressed individually in UD, and were collapsed into three coarse-grained correlations:
First, tense/aspect and negative auxiliaries are together represented by the \emph{aux} relation in UD.
Second, the relation between complementizers and adverbial subordinators with their complement clauses is represented by the \emph{mark} relation.
Third, both the verb-PP relation and the relation between adjectives and their standard of comparison is captured by the \emph{obl} relation.

@@

Here we discuss how our formalization of communicative efficiency relates to formalizations that have been proposed in the information-theoretic literature on language.
Across the literature, the core idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms, with some differences in the precise formalization of these two quantities \cite{ferreri2003least,ferrericancho2007global,kemp2012kinship,frank2012predicting,goodman2013knowledge,kao2014nonliteral,xu2014numeral,kirby2015compression,regier2015word,xu2016historical,futrell2017memory,zaslavsky2018efficient,bennett2018extremely, peloquin2019interactions,zaslavsky2019semantic}.

@@

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parseability.
While the computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$, we also examined word order predictions for grammars optimized for only parseability or only predictability, in order to tease apart predictions made by these two components.
As shown in Table~\ref{tab:all-predictions-1}, each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions.
%Indeed, the same holds for all of the 38 syntactic relations in UD  (Table~\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})
%We additionally report results with $\lambda = 0$, i.e., placing all weight on parseability.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.\footnote{Results from one of the preliminary experiments reported in Figure~\ref{table:corr-resu-previous} show that results are stable to small variation of $\lambda$: Essentially equivalent predictions are obtained for $\lambda=1.0$. While $\lambda=1.0$ is not a valid choice for communicative efficiency in general due to the possibility of collapse to a single utterance, our setting does not allow such a collapse, as the syntactic structure already determines which words are present in the sentence.}
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}), in which case not all correlations are predicted (Figure~\ref{fig:posterior}).
See also Figure~\ref{table:corr-resu-previous} for converging evidence from our preregistered preliminary experiments.
%Finally, the robustness of our results to small variations of $\lambda$ is confime by reporting results with $\lambda=1$ in Table~\ref{table:corr-resu-previous}.

@@

\begin{table}
\small{
\begin{center}
\begin{tabular}{l||ll|ll|ll|ll|ll|ll|llllll}
%	   for(dep in c("obj", "lifted_mark", "acl", "nmod", "obl")) {
		   Language 
		   &	\multicolumn{2}{|c|}{Objects} 
		   &	\multicolumn{2}{|c|}{Adpositions} 
		   &	\multicolumn{2}{|c|}{Compl.} 
		   &	\multicolumn{2}{|c|}{Rel.Cl.} 
		   &	\multicolumn{2}{|c|}{Genitive} 
		   &	\multicolumn{2}{|c|}{PP}  \\ \hline\hline
	\input{../results/grammars/comparison-table.tex}
\end{tabular}
\end{center}
}
\caption{Comparing grammars extracted from databases to linguistic judgments in the World Atlas of Linguistic Structures. For each of the six syntactic relation, the first column provides the ordered coded in the extracted grammar; the second column provides the order coded in WALS (DH for dependent-head, HD for head-dependent order). `?' indicates that WALS has no data.
$*$ indicates that WALS does not list a dominant order, which can mean that neither order is dominant in the language, or that insufficient data was available when compiling WALS \citep{dryer2011evidence}.
Finally, `--' indicates that the relation does not occur in the corpus.
}\label{tab:grammars-wals}
\end{table}

@@

Recall the definition of efficiency
\begin{equation}\label{eq:efficiency-recall}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where
\begin{equation}\label{eq:rpars}
	R_{Pars} := I[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)} 
\end{equation}
\begin{equation}
	R_{Pred} := - H[\utterance] = \sum_{u} p(u) \log p(u)
\end{equation}
where $t \sim \tree$ is a distribution over syntactic structures, and $u \sim p_{\mathcal{L}_\theta}(u|t)$ denotes the corresponding linearized sentences.

@@

The degree to which a parser with parameters $\phi$ succeeds in parsing a sentence $u$ with structure $t$ is\footnote{Note that, in the definition of $R_{Pars}$ (\ref{eq:rpars}), the term $p(t)$ is a constant independent of $\phi$ and the word order grammar $\mathcal{L}_\theta$; it can therefore be ignored in the optimization process.} 

@@

	R_{Pars}^{\phi}(u,t) =  \log p_\phi(t|u)

@@

Using this and Gibbs' inequality~\cite{cover2006elements}, we can rewrite Efficiency~(\ref{eq:efficiency-recall}), for a given grammar $\theta$, equivalently into the parseability and predictability achieved with the best parser and language models:

@@

	R_{\textit{Eff}}^{\theta} := \max_{\phi,\psi} R_{\textit{Eff}}^{\theta, \phi, \psi}

@@

where we have written
R_{\textit{Eff}}^{\theta, \phi, \psi} := \E_t \E_{u \sim p_\theta(u|t)} \left[R_{Pars}^{\phi}(u,t) + \lambda R_{Pred}^{\psi}(u)\right]

@@

In order to find an optimal grammar $\theta$, we thus need to compute 

@@

\argmax_\theta\	R_{\textit{Eff}}^{\theta} = \argmax_\theta\	\max_{\phi, \psi} R_{\textit{Eff}}^{\theta, \phi, \psi} 
%= \argmax_\theta\	\max_{\phi, \psi} \E_t \E_{u \sim p_\theta(u|t)} \left[R_{Pars}^{\phi}(u,t) + \lambda R_{Pred}^{\psi}(u)\right]

@@

where (\ref{eq:dtheta}) is derived using the \emph{score-function} or \emph{REINFORCE} theorem~\cite{williams1992simple}.

@@

We can therefore apply stochastic gradient descent to jointly optimize $\theta, \phi, \psi$:
In each optimization step, we sample a dependency tree $t$ from the treebank, then sample an ordering from the current setting of $\theta$ to obtain a linearized sentence ${\bf w} \sim p_{\theta}(\cdot|t)$.
Then we %update $\thetal$ with ordinary gradient descent, and $\thetad$ with the REINFORCE estimator~\ref{williams-simple-1992}.
do a gradient descent step using the estimator given by the expressions in the square brackets in (\ref{eq:dtheta}-\ref{eq:dpsi}).
